---
title: "Spotify_Final_Project"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
/-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
Title: Spotify Unwrapped
Group: Jonah Landas, Ryan Freidhoff

Questions: Can we profile the playlist to see exactly what the typical song looks like? What are the strongest correlations within these songs, and what exactly is the backbone of a playlist?

/-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
Introduction: The following project will be aimed at discovering correlations and unique aspects of a custom playlist we made on Spotify. This is of interest as we would like to see if our analysis can profile what type of listeners we are and possibly suggest how to look for new songs based on our profiling of the typical song on the playlist. We will begin by describing the data, cleaning and reformatting the data, and then using visualization tools to recognize trends and patterns within the data before summarizing our findings in the conclusion. The visualization and findings section of this report will comprise of visualizations and summaries of the information learned from each visual.

Data: Data is from a playlist I have made on Spotify. The data contains date each song was produced, song duration, artist name, album name, song name, if it is a one-hit-wonder (Y/N), how many Spotify plays the song has, type of release, number of monthly listeners the main artist has, and a calculated column we added that is a ratio of the number of plays divided by the number of monthly listeners in an effort to normalize the values.


We will begin by loading in the .csv file. The chunk of code below reads the .csv file and creates the dataframe 'spotify' that we will be using throughout this project. Additionally, the column `Release.Date` is turned into type date and the `Duration` column is also transformed into minutes and seconds using lubridate.

```{r}
  library(rvest)
  library(lubridate)
  library(hms)

  csv <- read.csv("DS 202 Final Project - Data(Sheet1).csv")
  spotify <- as.data.frame(csv)
  spotify$`Release.Date` <- lubridate::mdy(spotify$`Release.Date`)
  spotify$Duration <- lubridate::ms(spotify$Duration)
  head(spotify)
```

The first piece of investigation will be to dive into the number of plays each song has compared
to the number of monthly listeners the main artist has. We will add a column that gives the ratio
between the number of plays and the number of monthly listeners in the chunk below. We will name
this new column `Plays.Listener.Ratio`.

```{r}
  library(dplyr)
  library(tidyr)

  #This ratio of the plays by the monthly listeners (fan base) helps us normalize the values
  spotify$Plays.Listener.Ratio <- spotify$Plays / spotify$`Monthly.Listeners`
  
  #avgMonthlyListeners will be used later in some of the graphing
  avgMonthlyListeners <- mean(spotify$`Monthly.Listeners`)
  #avgMonthlyListeners
  
  #avgRatio will be used later for some of the graphing
  avgRatio <- mean(spotify$Plays.Listener.Ratio)
  #avgRatio
  
  #avgPlays will be used later for some of the graphing
  avgPlays <- mean(spotify$Plays)
  #avgPlays
```

```{r}
  library(ggplot2)
  
  #This is our initial visualization of the plays by the monthly listeners in the form of a scatter plot
  plot1 <- ggplot(data = spotify, aes(x = `Plays`, y = `Monthly.Listeners`)) +
  geom_point() 
  
  plot1
```

The above graph shows the initial investigation to determine if plays and the number of monthly listeners are highly correlated. Based on the scatter plot, we can see that there is a very weak positive correlation between the two variables.

```{r}
  library(ggplot2)

  #Our second layer of visualization adds the avgPlays and avgMonthlyListeners to the graph
  plot2 <- ggplot(data = spotify, aes(x = `Plays`, y = `Monthly.Listeners`)) +
  geom_point() +
    geom_hline(aes(yintercept = avgMonthlyListeners)) +
    geom_vline(aes(xintercept = avgPlays)) +
    geom_text(aes(5e8, y= avgMonthlyListeners,label = "y = 15,281,061", vjust = -0.5)) +
    geom_text(aes(x= avgPlays, y=4e7,label = "x = 235,524,797", vjust = 1.5, angle = 90))
  
  plot2
```

We further the investigation by adding horizontal and vertical lines at the average number of monthly listeners and the average number of plays, respectively. This was done in an effort to see if there is a 'average' song based on these metrics alone. There are no songs particularly close to the intersection of the averages, so we will investigate if normalizing the number of plays by creating the previously mentioned plays to monthly listeners ratio will yield any interesting results.

```{r}
  library(ggplot2)

  #We now use the normalized ratio calculated early to see the typical tatio of plays to fan base in this histogram
  plot3 <- ggplot(data = spotify, aes(x = Plays.Listener.Ratio)) +
    geom_histogram()

  plot3
```

To begin, we create a histogram containing the plays to monthly listeners ratio. As we can see, the distribution is unimodal skewed right with the average ratio being approximately 20. Our next move will be to test if either of our categorical variables are heavily dependent with the ratio.


```{r}
  library(ggplot2)

  #We now add the get_density function to see the relative density of each ratio in terms of one hit wonder column
  plot5 <- ggplot(data = spotify, aes(x = Plays.Listener.Ratio,y= after_stat(density) ,fill= `One.Hit.Wonder..Y.N.`, color= `One.Hit.Wonder..Y.N.`)) +
  geom_histogram(binwidth = 3) +
  geom_density(alpha=.5, fill= "black")
  
  plot5
```

From the above graph, we can see that not only do one hit wonders have a much larger average plays to monthly listeners ratio, but the range is also far greater for the one hit wonder songs than the tight spread of the songs that are not one hit wonders. We used the geom_density fucntion to show the general makeup of each category, which emphasized the difference in the mean ratio of the one hit wonders compared to the songs that are not one hit wonders.


```{r}
  library(ggplot2)

  #We now add the get_density function to see the relative density of each ratio in terms of release type column
  plot6 <- ggplot(data = spotify, aes(x = Plays.Listener.Ratio,y= after_stat(density) ,fill= `Type.of.Release`, color= `Type.of.Release`)) +
  geom_histogram(binwidth = 15) +
  geom_density(alpha=0.5, fill="black")
  
  plot6
```

The above graph shows how the relative density of plays to monthly listeners ratio differs based on the type of release of the song. We can see that although there are fewer singles than albums, the overlay that shows a smooth function of the relative density shows that the singles and albums are nearly identical in terms of the plays to fan base ratio. Additionally, the songs released on EPs are mainly evenly accross a ratio less than 1 all the way to about 150. 

These findings warranted further research as to what an EP is and why songs are released on EPs. After researching EPs, we found that EPs are often used by smaller artists as a cheap alternative to an album. EPs typically contain 3-6 songs, compared to albums that usually contain more than 10 songs. Based on the graph, we can see that the average ratio for the songs released on EPs is about 100, meaning that those songs are most likely also one hit wonders, backing the claim that smaller artists on this playlist are more likely to have a one hit wonder and as a result a larger plays to monthly listeners ratio due in part by the smaller fan base.

```{r}
  library(ggplot2)

  #We can now investigate whether the release date is correlated with the number of plays
  ggplot(data = spotify, aes(x = `Release.Date`, y = `Plays`)) +
    geom_point()
```
From the above scatter plot, we see that there is not a correlation between the number of plays and the release date. We will now look into if the plays to number of monthly listeners ratio is at all correlated to the song duration.

```{r}
  library(ggplot2)

  #We can now investigate whether the release date is correlated with the number of plays
  ggplot(data = spotify, aes(x = `Plays.Listener.Ratio`, y = `Duration`)) +
    geom_point()
```

The above graph shows the relationship between the plays listener ratio and the song duration in seconds. We can see a very slight positive correlation between the two variables. The relationship also appears slightly linear in nature, with one major outlier in the y direction.


```{r}
  library(modeest)
  library(lubridate)

  summary(spotify$Release.Date)
  #The summary shows us that the average release date and the median release date are both in 2017 which agrees with the above scatter plot. The oldest song on this playlist is from 2001 and the most recent is from late 2020.
  mlv(month(spotify$Release.Date), method = "mfv")
  mlv(wday(spotify$Release.Date), method = "mfv")
  summary(spotify$Duration) #Shows the average duration of all songs
```

Based on the above code chunk, we can see that songs on this playlist are most likely to be produced in February, and the most often day of the week that the songs are released is Friday. We also see that the average song duration is 3 minutes and 23 seconds with the shortest song being only 1 minute and 57 seconds and the longest song being 7 minutes and 48 seconds, showing just how large the range of the duration is.

```{r}
  library(corrplot)
  library(lubridate)

  spotifyWork <- spotify
  spotifyWork$Seconds <- as.integer(seconds(spotifyWork$Duration))
  spotifyWork$Year <- as.integer(year(spotifyWork$Release.Date))
  spotifyNumeric <- spotifyWork[c(1, 9, 10, 11, 12, 13)]
  
  head(spotifyNumeric)
  cor_data = cor(spotifyNumeric)
  corrplot(cor_data)
```

The above code chunk shows how we were able to create a correlation matrix of the data. We had to first isolate only the numeric columns of the data. We called this new data frame `spotifyNumeric`. We then created the actual matrix using the corrplot package. A few things to note from the matrix are that the `Plays` and `Plays.Listener.Ratio` columns are completely uncorrelated. The most positively correlated columns are the `Plays` and the `Monthly.Listeners` columns. The most negatively correlated columns are the `Plays.Listeners.Ratio` and `Monthly.Listeners` columns.

```{r}
  library(tidyverse)
  library(tidyr)

  length(unique(spotify$Artist))
  spotify2 <- separate(data = spotify, col = `Artist`, into = c("Artist1", "Artist2", "Artist3", "Artist4"), sep= ",")
  pivoted <- spotify2 %>% pivot_longer(cols = c("Artist1", "Artist2", "Artist3", "Artist4"), 
                                      names_to = "Column Number", values_to = "Artist Name")
  #pivoted
  artistCount <- pivoted %>% count(`Artist Name`, sort= TRUE)
  head(artistCount)
  #spotify2
```


The above chunk of code was used to find which artists appeared the most on the playlist. We had to first reformat the data using pivot_longer to account for songs with multiple artists. We had to create 4 columns due to one song having 4 artists on it. From this, we were able to determine the top 5 most common artists in the playlist, blackbear appearing the most with 9 songs. There were a total of 84 unique artists across the 100 songs.

```{r}

  #This shows the most played songs
  popularSongs <- spotify[order(spotify$Plays, decreasing = TRUE),]
  head(popularSongs)
```

Issues Along the Way:
Obtaining the data required a few hours of time due to the data set being custom. The solution was to be persistent and change some of the columns originally planned on being in the data set.

read_csv was being used originally and was working. Halfway through analysis, errors were thrown regarding read_csv() being an unknown function. The solution was using read.csv() instead and then updating the column names.

Date Released column format in .csv file was not compatible with lubridate. The solution was reformatting the excel worksheet to a different display of the dates and updating the .csv file version.

Some of the artists had characters that are not allowed in r in their name. The solution was to alter the name in the .csv file.



Conclusion: Based on the above analysis, we can see that the typical song in this playlist is 3:23 in duration, released on a Friday and most likely released in 2017. Additionally, we have that the most prevalent artists in this playlist are blackbear (9 songs), Mac Miller (6 songs), Aminé (4 songs), Drake (3 songs) and Kota the Friend (3 songs). The top 5 songs in terms of Spotify plays are "Redbone", "Daddy Issues", "Falling Down - Bonus Track", "What You Know Bout Love", "Give Me Everything (feat. Nayer)" and "Laugh Now Cry Later (feat. Lil Durk)". Overall, R&B was the most popular genre by far. We hypothesize that these findings are highly personalized as this playlist was all custom and actually took hours to originally create. We believe that each playlist creator will have a unique 'playlist profile' accentuated by the histograms containing the plays by monthly listeners ratio. We also firmly believe that the amount of time spent creating the playlist has a strong correlation with how personalized the fore mentioned 'playlist profile' is. We decided to to display our conclusion using the Spotify Wrapped template that Spotify releases each year for each of its listeners.