---
title: "Spotify_Final_Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Title: Spotify Data Final Project
Group: Jonah Landas, Ryan Freidhoff
Data: Data is from a playlist I have made on Spotify. The data contains year each song was produced, duration, artist name, album name, song name, if it is a one-hit-wonder (Y/N), popularity rating, type of release, and genre.
Questions: Can we profile the playlist to see exactly what the typical song looks like? What are the strongest correlations within these songs, and what exactly is the backbone of a playlist?

```{r}
  library(rvest)

  spotify <- read_csv("DS 202 Final Project - Data(Sheet1).csv")
  head(spotify)
```

The first piece of investigation will be to dive into the number of plays each song has compared
to the number of monthly listeners the main artist has. We will add a column that gives the ratio
between the number of plays and the number of monthly listeners in the chunk below. We will name
this new column `PlaysListenerRatio`.

```{r}
  library(dplyr)

  spotify$PlaysListenerRatio <- (spotify$Plays / spotify$`Monthly Listeners`)
  spotify
```
